A crop modelling strategy to improve cacao quality and productivity

### 2021-12-15 paperModel/paperCacaoModel_round1.tex
round1 corrections

%=================================================================
% Add packages and commands here. The following packages are loaded in our class file: fontenc, inputenc, calc, indentfirst, fancyhdr, graphicx, epstopdf, lastpage, ifthen, lineno, float, amsmath, setspace, enumitem, mathpazo, booktabs, titlesec, etoolbox, tabto, xcolor, soul, multirow, microtype, tikz, totcount, changepage, paracol, attrib, upgreek, cleveref, amsthm, hyphenat, natbib, hyperref, footmisc, url, geometry, newfloat, caption
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
%=================================================================
%% Please use the following mathematics environments: Theorem, Lemma, Corollary, Proposition, characterisation, Property, Problem, Example, ExamplesandDefinitions, Hypothesis, Remark, Definition, Notation, Assumption
%% For proofs, please use the proof environment (the amsthm package is loaded by the MDPI class).

@@

\abstract{Cacao production systems in Latin America have a high importance over social and economic development, facing  the fight against hunger and poverty. Although Colombian cacao has the potential to be in the high value markets for fine flavour, the lack of expert support as well as the use of traditional, and often times suboptimal, technologies makes cocoa production negligibly. For example, traditionally cacao harvest takes place at 5 or 6 months after flowering, other environmental parameters that have more association with pod maturation speed are not taken into account.  Cacao fruits development can be considered as the result of a number of physiological and morphological processes that can be described by mathematical relationships even under uncontrolled environments. In this context, crop models are useful tools to simulate and predict crop development over time and under multiple environmental conditions. Since, harvesting at the right time can yield high quality cacao, we parametrised a crop model to predict the best time for harvest cocoa fruits in Colombia. Our aim was to develop a practical tool that supports cacao farmers in the production of high quality cacao. When comparing simulated and observed data, our results showed an RRMSE of 7.2\% for the yield prediction, while the simulated harvest date varied between +/- 2 to 20 days depending on the temperature variations of the year between regions. This crop model contributed to understand and predict the phenology of cacao fruits for two key cultivars ICS95 y CCN51.}

@@

Cacao ( \textit{Theobroma cacao }L. ) is an important worldwide perennial tropical crop endemic to the South American rainforests \citep{zuidema2005, motamayor2002, argout2011, Rodriguez2019}. Cacao plant is a member of the Malvaceae (formerly Sterculiaceae)  botanical family such as  cotton \textit{ Gossypium hirstium} \citep{Nix2017cotton}. Cotton has been is modeled in SIMPLE model \citep{Zao2019simple}. Cacao is grown for its fruits, known as cacao pods \citep{ Niemenak2010, suarez2021}. Only the 5\% of the world cocoa yield is destined for Fine-cocoa production due to the low productivity through  the traditional crop management \citep{argout2011}.  In Colombia, cocoa  is  traditionally  consumed  as  a  beverage. It is one of the crops promoted by the Colombian government in the social and agricultural development  programs aimed at favouring peace in post-conflict regions \citep{Rodriguez2019, Abbott2019}. This crop is grown by approximately 52.000 families \citep{Gutierrez2020} and 98\% of production being carried out by small and medium-sized producers \citep{Garcia2014, Escobar2020}. Colombia registered an increase of 3.750 tons in production in 2020 compared to the previous year \citep{lamos2020}.

Although Colombian cocoa has the potential to be in the high value markets for fine flavour \citep{Escobar2020}, the lack of expert support as well as the use of traditional, and often times suboptimal, technologies makes cocoa production negligibly. For example, the farmer practice of empirically harvesting from 5 to 6 months or 180 Days After Flowering  (DAF) date, produced a mix of quality of cacao beans that are then fermented. This practice produces heterogeneous characteristics between each fermentation batch which potentially diminish the quality of the final cocoa product \citep{Escobar2021}. To identify the best moment to harvest is important to consider physiological responses affected by climate variables such as rain, solar radiation and wind.  Therefore,  for cacao production in Colombia, physiological simulation models may be valuable to identify the best moment to harvest cocoa fruits considering variable weather conditions, soil types and cultivar specifications.

Crop models represent a quantitative assumption of plant growth depending on sunlight interception efficiency values and climate data supported by a large amount of empirical and ground data. \citep{Reynolds2018}. Physiological crop models have shown to be very useful tool for provided agronomical advices and improvements of the cropping systems of annual crops mainly. Recently crop modelling studies are focusing on  perennial crops  production \citep{zuidema2005, Zao2019simple, Bai2020, Romero2021}. However, the information reported  is be less than for annual crops due to the lack of field data available , relatively high research costs and the difficulties of accumulated errors in long-term simulations \citep{zuidema2005}. For cacao there are  approaches  to predict yield mainly  using algorithms of machine learning \citep{lamos2020} and just one mechanistic model simulates physiological cocoa performance  "SUCROS-cocoa" \citep{zuidema2005}. This crop model calculates light interception, photosynthesis, maintenance respiration,
evapotranspiration, biomass production and cocoa yield. It can be parametrised having data on cocoa physiology and morphology \citep{zuidema2005}. However,  there is not specific cocoa physiology data available from small and medium-sized producers. Thus, we adapted the simple generic crop model (SIMPLE) that could be easily modified for any crop to simulate development, crop growth and yield using few parameters such as weather and cultivar specification \citep{Zao2019simple}.

In this paper, we present a physiological parametrisation of the SIMPLE crop model \citep{Zao2019simple} for cocoa to predict best harvest time and overall yield in the framework of KOCOLATL project, which is a collaborative work between Colombian partners (researches, farmers and organizations such as Fedecacao and BIOS) and researchers in NIAB from United Kingdom. We parametrised this crop model  because it had already been successfully fitted to other tree crops in south America. The model simulates crop development, growth and yield, and predicts the maturation day when the fruit is likely to be ready for harvest. Our aim was to develop a practical tool that supports cacao farmers in the production of high quality cocoa. Thus, making Colombian cacao more competitive in the Fine-cocoa market.

\subsection{Floral Phenology of Cocoa}

@@

@@

The cocoa model performance was evaluated by comparing simulated values cocoa yield with those reported by Fedecacao from cocoa plantations, using the statistical indice of Relative Root Mean Square Error (RRMSE)  in the equation \ref{equ:RRMSE}, where n is the total number of observations, Y$_{i}$ corresponds to observed value from field and X$_{i}$ is the predict value from the model.\\

@@

\subsection{Weather Conditions Over Flowering Time }
We analysed and compared weather patterns linked our five geographical regions (Fig.\ref{fig:yield}, a) and across 2018 to 2020 . Looking at the data, the regions of Santander and Caldas had the biggest variability and the maximum of solar radiation values over 20 MJ m$^{2}$day$^{-1}$ . In contrast , the regions of Cali and Arauca presented the lowest values of PAR below 5 MJ m$^{2}$day$^{-1}$. (Fig.\ref{fig:temp}, a).  Even though, Cali and Santander had contrasting PAR conditions, these regions presented  have similar temperature during 2020 (Fig.\ref{fig:temp}, b). The temperature ranges from 16 to 28 $^\circ$C and it is relatively constant for each region. However, Arauca presented the biggest variability with hotter months during the first half of the year 2019 and 2020.  Apartado was found as the hottest region studied with a relative constant temperature of 26 $^\circ$C and Caldas as the coldest site with 16 $^\circ$C.  Precipitation in Colombia is presented in two seasons per year from February to April and from October to November, while the relative humidity remain constant over 80\% (Fig.\ref{fig:temp},c). In general, the coldest regions tested (Caldas and Santander) had the maximum values of solar radiation available for photosynthesis.

Figure \ref{fig:heat} shows the pearson correlation to study the weather data of the flowering time over the months of flowering  (monthF), month of harvest (monthH) and their final yield (fruit\_kg). The results showed that thermal time T$_{b}$ (ttb) is correlated (P = 0.52) with daily average temperature and maximum temperature (TMAX) and  temperature minimum (TMIN) and Dew Frost Point at 2 meters (T2MDEW) with  a correlation coefficient of 0.60. The wind (WS2M) is correlated with ttb with 0.57. However, less clear correlations were found of monthF  with T2MDEW, relative humidity (RH), WS2M and rain. Although, farmers stated that the number of flowers pollinated decrease by months where wind and rain are high (personal communication with Fedecacao), our analysis  (Fig. \ref{fig:heat}) could not show a correlation between flower shedding and high wind or rain because information on number of  flowers at FD was not part of the field data collected.

@@

Thermal time characterisation was made considering 180 DAF for each location. The boxplot in the figure \ref{fig:ttbox} shows the data distribution where boxes indicate the range of the central 50\% of the total data per region, the central line in the box is marking the median value and lines draw out from each box mean the range of the remaining data. Therefore, this boxplot shows differences between locations as was expected following the tendencies of the temperature of the figure \ref{fig:temp}, (b), where Aparatado, Arauca and Cali had greater temperatures than Santander and Caldas.

@@

The RRSME has been used before to evaluated other crop model simulations \citep{Zao2019simple, Bai2020}. Our results of cocoa yield simulation (kg ha$^{-1}$ per year) and validation (table \ref{tab:reparam}) achieved a final RRMSE of 7.2\%  fit between simulated and observed data (fig. \ref{fig:m1}, c).  Individual errors per region are presented in table \ref{tab:error}, where the best fit of the calibration model for yield prediction  was for crops in  Apartado and the highest error was calculated for Caldas crops.  The model responded to the variations of temperature and solar radiation. Therefore, the highest  yield values simulated were obtained for Arauca over 4000 kg ha$^{-1}$, followed by Apartado Santander with yields over 2000 kg ha$^{-1}$.  The lowest yield was simulated for the Caldas region with less of 1000 kg ha$^{-1}$ . Final yield in the model was calculated as the product of biomass of aerial part and harvest index (HI) \citep{Zao2019simple, Amir1991}, where the HI is similar to the CropSyst \citep{STOCKLE2003} and AquaCrop \cite{Steduto2009}.

@@

Cacao is one of the most economically important crops in Colombia. In 2016 Colombia produced a historical of 56K tons, up from 35K ten years ago, and over the same period, imports decreased from 12.8K to 488 tons (FAOStat). Cacao has been fundamental in the transformation of the rural economy dependent on illicit crops. As such, Cacao is called the "peace crop", nominated by  Colombia's President and 2016 nobel peace prize winner, Juan Manuel Santos, and by the President of the Association of Cacao Growers (Fedecacao). Although Colombian cacao has the potential to be in the high value markets for fine flavour, the lack of expert support as well as the use of traditional, and often times suboptimal, technologies makes cocoa production negligibly. This study reports the cacao model which was calibrated for five agroclimatic regions in Colombia. The purpose of this model was to help farmers achive  higher quality cacao beans by providing them with a tool that not only predicts optimal harvest date but also estimated yield and biomass.

@@

Our analysis of weather variables (Fig. \ref{fig:heat}) could not show a correlation between flower shedding and high wind or rain because information on flowers was not part of the field data collected. However, farmers stated that the number of flowers pollinated decrease by months where wind and rain are high. High winds can affect the availability of tiny flies pollinators from Diptera order and from the families of of the biting midges \textit{Ceratopogonidae},  genus  \textit{Forcipomyia} \citep{Saunders1959, kaufmann1975, sotomayor2020} to reach the cacao flowers. However, the stability of cacao flowers is influenced by seasonal wheather conditions (abiotic) and pollination (biotic) \citep{Frimpong2014}. Therefore, pollinator population should be coincidence with the phenology of the flowering cacao trees \citep{Young1983, Young2012}. Flower opening is very well synchronised between the cohorts of mature flowers opening each night \citep{Niemenak2010}. The flowers open at almost exactly the same time and rate, irrespective of their position on the trunk. Thus, unfertilised flowers abscise from the trunk approximately 1 day after flower opening  \citep{Niemenak2010}. Hence more than 90\% of unpollinated flowers fall or abscised within 32 hours after  anthesis \citep{Aneja1999}. Abscission processed of  flowers are mainly controlled  by  three  hormones: auxin, ethylene, and abscisic acid (ABA) \citep{Aneja1999}. Ethylene generally promotes abscission because it may  inhibit the transport of auxin from the leaf blade, which allow the action of ABA to promove the fall of flowers \citep{Beyer1975}. In general, environmental conditions can also  stimulate   to  a  decrease  the  auxin/ethylene relationship \citep{Aneja1999}.

When analysing the data from the Santanter region, we showed that the number of successful flowers pollinated to produce final yield could be affected by rain, TMAX and wind. Nevertheless, a better field data tracing flowers development is essential to understand if it is a mechanical or physiological effect.  \cite{Frimpong2009, Frimpong2011} indicated that the numbers of cocoa pollinators were reduced during the dry season, but increased in the wet season. This, could be due to midges need a moist environment to develop \citep{Frimpong2014}, which is difficult during the dry season as cocoa leaves create a dried ground mat \citep{Frimpong2009, Frimpong2014}. Moreover, the lack of water during dry seasons may reduce the nutrients uptake provoking the massive flower drops \citep{Vaughton2017}.  For future studies, wind could be included in the cocoa model as an input to simulate this mechanical effects over number of flowers. In general, field data regarding counting flowers pollinated by month should be better reported for the region here studied.

\subsection{Thermal Time for Harvest Day Predictions}
We define the thermal time required to harvest cocoa pods for five Colombian regions as maturation of the fruit is related with temperature during the growth cycle \citep{lopez2018}. Previous studies have calculated thermal time for  different cocoa cultivar in Brazil and Ghana \cite{Daymond2008}. They also confirm that the fruit maturation time decrease in with an increase in temperature as was presented on others researches \citep{Alvim1974, End1991, Daymond2008}. the effects of temperature and solar radiation on fruit growth and development was previously studied by \cite{Daymond2008}, showing that crop under higher temperatures thought the crop cycle induce greater fruit losses because of physiological maturation (cherelle wilt). When the fruit is mature, seeds are able to germinate inside the pod before the harvest day \citep{lopez2018}.

@@

\subsection{Cocoa Crop Model Simulations}
Although cocoa is a relevant crop and there is an extensive agronomic literature, there is only one physiological crop model specific for cocoa so far. The (SUCROS-Cocoa) developed by \cite{zuidema2005}. However, the code was not easy available for adaptations. In contrast, the SIMPLE model has an open code in R, which we could adapt such a model would be very useful to compare yields and predict harvest date in different climates. As the harvest day was predicted from the FD (Fig. \ref{fig:dayH}, consequently, biomass production and fSolar presented a crop cycle shorter than 180 DAF (Fig. \ref{fig:m1}, a and b). These simulations are coincident with results presented by \cite{lopez2018}, where physiological maturity of coca pod varies from 140 to 162 DAF.  Our results showed how the harvest day can vary depending on the accumulate temperature during each especific crop cycle simulated (Fig. \ref{fig:dayH}).

@@

Yield prediction presented a RRMSE values 7.2\%, which were significant lower than those  presented for other crops using the SIMPLE model which reported and RRMSE of 24.4\% \citep{Zao2019simple}. Resulting in reliable approach for cocoa yield prediction. In general, these results may help to improve the quality of cocoa seed considering the moment to harvest can be variable depending on weather changes.

@@

Cocoa fruit development for harvest in the right time depend on whether conditions and principles of crop physiology and flower phenology.  This was common for the five regions. Thermal time characterisation range from 1200 to 3000 days $^\circ$C, with a T$_{b}$ of 10 $^\circ$C for the fruit development.  The cocoa model allowed to predict the harvest date with better precision than only considering days by calendar. Thus, the crop cycle of cocoa for harvest  should be shorter than 180 days after flowering. These results confirm the potential of the Crop Simulation Model approaches  for tropical crop in Latin America.

>>>>

%=================================================================
% Add packages and commands here. The following packages are loaded in our class file: fontenc, inputenc, calc, indentfirst, fancyhdr, graphicx, epstopdf, lastpage, ifthen, lineno, float, amsmath, setspace, enumitem, mathpazo, booktabs, titlesec, etoolbox, tabto, xcolor, soul, multirow, microtype, tikz, totcount, changepage, paracol, attrib, upgreek, cleveref, amsthm, hyphenat, natbib, hyperref, footmisc, url, geometry, newfloat, caption
\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\usepackage[spanish,english]{babel}
%=================================================================
%% Please use the following mathematics environments: Theorem, Lemma, Corollary, Proposition, characterisation, Property, Problem, Example, ExamplesandDefinitions, Hypothesis, Remark, Definition, Notation, Assumption
%% For proofs, please use the proof environment (the amsthm package is loaded by the MDPI class).

@@

\abstract{Cacao production systems in Latin America have a high importance over social and economic development, facing  the fight against hunger and poverty. Although Colombian cacao has the potential to be in the high value markets for fine flavour, the lack of expert support as well as the use of traditional, and often times suboptimal, technologies makes cocoa production negligibly. \textcolor{blue}{Traditionally cacao harvest takes place at 5 or 6 months  by calendar after flowering, the problem is that  the seeds are often  unripe or  over ripe germinating inside the pod afecting the quality of the final cocoa products}. Other environmental parameters that have more association with pod maturation speed are not taken into account.  Cacao fruits development can be considered as the result of a number of physiological and morphological processes that can be described by mathematical relationships even under uncontrolled environments. In this context, crop models are useful tools to simulate and predict crop development over time and under multiple environmental conditions. Since, harvesting at the right time can yield high quality cacao, we parametrised a crop model to predict the best time for harvest cocoa fruits in Colombia. . Results can be a practical tool that supports cacao farmers in the production of high quality cocoa.  When comparing simulated and observed data, our results showed an RRMSE of 7.2\% for the yield prediction, while the simulated harvest date varied between +/- 2 to 20 days depending on the temperature variations of the year between regions. This crop model contributed to understand and predict the phenology of cacao fruits for two key cultivars ICS95 y CCN51. \textcolor{blue}{The aim of this study was to calibrate a crop model that simulates cacao crop development and yield, and predicts the maturation day when the fruit is likely to be ready for harvest.}}

@@

Cacao ( \textit{Theobroma cacao }L. ) is an important worldwide perennial tropical crop endemic to the South American rainforests \citep{zuidema2005, motamayor2002, argout2011, Rodriguez2019}. Cacao plant is a member of the Malvaceae (formerly Sterculiaceae)  botanical family such as  cotton \textit{ Gossypium hirstium} \citep{Nix2017cotton}. Cotton has been is modelled in SIMPLE model \citep{Zao2019simple}. Cacao is grown for its fruits, known as cacao pods \citep{ Niemenak2010, suarez2021}. Only the 5\% of the world cocoa yield is destined for Fine-cocoa production due to the low productivity through  the traditional crop management \citep{argout2011}.  In Colombia, cocoa  is  traditionally  consumed  as  a  beverage. It is one of the crops promoted by the Colombian government in the social and agricultural development  programs aimed at favouring peace in post-conflict regions \citep{Rodriguez2019, Abbott2019}. This crop is grown by approximately 52.000 families \citep{Gutierrez2020} and 98\% of production being carried out by small and medium-sized producers \citep{Garcia2014, Escobar2020}. Colombia registered an increase of 3.750 tons in production in 2020 compared to the previous year \citep{lamos2020}.

Although Colombian cocoa has the potential to be in the high value markets for fine flavour \citep{Escobar2020}, the lack of expert support as well as the use of traditional, and often times suboptimal, technologies makes cocoa production negligibly. The farmer practice of empirically harvesting from 5 to 6 months or 180 Days After Flowering  (DAF) date, produced a mix of quality of cacao beans that are then fermented. \textcolor{blue}{The problem is that  the seeds are often  unripe or  over ripe germinating inside the pod affecting the cocoa quality}. This practice produces heterogeneous characteristics between each fermentation batch which potentially diminish the quality of the final cocoa product \citep{Escobar2021}.  \textcolor{blue}{Therefore,  It is important to identify the best moment to harvest  considering physiological responses affected by  soil types , cultivar specifications and weather variables such as rain, solar radiation and wind.  Thus,  for cacao production in Colombia, physiological simulation models may be valuable to predict the right moment to harvest cocoa fruits.}

Crop models represent a quantitative assumption of plant growth depending on sunlight interception efficiency values and climate data supported by a large amount of empirical and ground data. \citep{Reynolds2018}. Physiological crop models have shown to be very useful tool for provided agronomical advices and improvements of the cropping systems of annual crops mainly. Recently crop modelling studies are focusing on  perennial crops  production \citep{zuidema2005, Zao2019simple, Bai2020, Romero2021}. However, the information reported  is less than for annual crops due to the lack of field data available , relatively high research costs and the difficulties of accumulated errors in long-term simulations \citep{zuidema2005}. For cacao there are  approaches  to predict yield mainly  using algorithms of machine learning \citep{lamos2020} and just one mechanistic model simulates physiological cocoa performance  "SUCROS-cocoa" \citep{zuidema2005}. This crop model calculates light interception, photosynthesis, maintenance respiration,
evapotranspiration, biomass production and cocoa yield. It can be parametrised having data on cocoa physiology and morphology \citep{zuidema2005}. However,  there is not specific cocoa physiology data available from small and medium-sized producers. Thus, we adapted the simple generic crop model (SIMPLE) that could be easily modified for any crop to simulate development, crop growth and yield using few parameters such as weather and cultivar specification \citep{Zao2019simple}.

In this paper, we present a physiological parametrisation of the SIMPLE crop model \citep{Zao2019simple} for cocoa to predict best harvest time and overall yield in the framework of KOCOLATL project, which is a collaborative work between Colombian partners (researches, farmers and organizations such as Fedecacao and BIOS) and researchers in NIAB from United Kingdom. We parametrised this crop model  because it had already been successfully fitted to other tree crops in south America.  Our aim was to calibrate a crop model that simulates crop development, growth and yield, and predicts the maturation day when the fruit is likely to be ready for harvest. Results can be a practical tool that supports cacao farmers in the production of high quality cocoa. Thus, making Colombian cacao more competitive in the Fine-cocoa market.

\subsection{Floral Phenology of Cacao}

@@

\subsection{Weather Conditions }
We analysed and compared weather patterns linked our five geographical regions (Fig.\ref{fig:yield}, a) and across 2018 to 2020 . Looking at the data, the regions of Santander and Caldas had the biggest variability and the maximum of solar radiation values over 20 MJ m$^{2}$day$^{-1}$ . In contrast , the regions of Cali and Arauca presented the lowest values of PAR below 5 MJ m$^{2}$day$^{-1}$. (Fig.\ref{fig:temp}, a).  Even though, Cali and Santander had contrasting PAR conditions, these regions presented  have similar temperature during 2020 (Fig.\ref{fig:temp}, b). The temperature ranges from 16 to 28 $^\circ$C and it is relatively constant for each region. However, Arauca presented the biggest variability with hotter months during the first half of the year 2019 and 2020.  Apartado was found as the hottest region studied with a relative constant temperature of 26 $^\circ$C and Caldas as the coldest site with 16 $^\circ$C.  Precipitation in Colombia is presented in two seasons per year from February to April and from October to November, while the relative humidity remain constant over 80\% (Fig.\ref{fig:temp},c). In general, the coldest regions tested (Caldas and Santander) had the maximum values of solar radiation available for photosynthesis.

@@

The cocoa model performance was evaluated by comparing simulated values cocoa yield with those reported by Fedecacao from cocoa plantations, using the statistical index of Relative Root Mean Square Error (RRMSE)  in the equation \ref{equ:RRMSE}, where n is the total number of observations, Y$_{i}$ corresponds to observed value from field and X$_{i}$ is the predict value from the model.\\

@@

\subsection{Weather Conditions Over Flowering Time }
Figure \ref{fig:heat} shows the Pearson correlation to study the weather data of the flowering time over the months of flowering  (monthF), month of harvest (monthH) and their final yield (fruit\_kg). The results showed that thermal time T$_{b}$ (ttb) is correlated (P = 0.52) with daily average temperature and maximum temperature (TMAX) and  temperature minimum (TMIN) and Dew Frost Point at 2 meters (T2MDEW) with  a correlation coefficient of 0.60. The wind (WS2M) is correlated with ttb with 0.57. However, less clear correlations were found of monthF  with T2MDEW, relative humidity (RH), WS2M and rain. Although, farmers stated that the number of flowers pollinated decrease by months where wind and rain are high (personal communication with Fedecacao), our analysis  (Fig. \ref{fig:heat}) could not show a correlation between flower shedding and high wind or rain because information on number of  flowers at FD was not part of the field data collected.

@@

Thermal time characterisation was made considering 180 DAF for each location. The box-plot in the figure \ref{fig:ttbox} shows the data distribution where boxes indicate the range of the central 50\% of the total data per region, the central line in the box is marking the median value and lines draw out from each box mean the range of the remaining data. Therefore, this boxplot shows differences between locations as was expected following the tendencies of the temperature of the figure \ref{fig:temp}, (b), where Aparatado, Arauca and Cali had greater temperatures than Santander and Caldas.

@@

The RRSME has been used before to evaluated other crop model simulations \citep{Zao2019simple, Bai2020}. Our results of cocoa yield simulation (kg ha$^{-1}$ per year) and validation (table \ref{tab:reparam}) achieved a final RRMSE of 7.2\%  fit between simulated and observed data (fig. \ref{fig:m1}, c). \textcolor{blue}{ The low RRMSE  indicates the simplicity of the Cacao model to simulated cacao  yield. Although, the original SIMPLE model can be used to asses heat and water stress, our cacao model calibration  did  not accounted biotic (pest and diseases) and abiotic (heat or water stress) factors that could have significant effects on the cacao production.}

Individual errors per region are presented in table \ref{tab:error}, where the best fit of the calibration model for yield prediction  was for crops in  Apartado and the highest error was calculated for Caldas crops.  
 The model responded to the variations of temperature and solar radiation. Therefore, the highest  yield values simulated were obtained for Arauca over 4000 kg ha$^{-1}$, followed by Apartado Santander with yields over 2000 kg ha$^{-1}$.  The lowest yield was simulated for the Caldas region with less of 1000 kg ha$^{-1}$ . Final yield in the model was calculated as the product of biomass of aerial part and harvest index (HI) \citep{Zao2019simple, Amir1991}, where the HI is similar to the CropSyst \citep{STOCKLE2003} and AquaCrop \cite{Steduto2009}.

@@

Cacao is one of the most economically important crops in Colombia. In 2016 Colombia produced a historical of 56K tons, up from 35K ten years ago, and over the same period, imports decreased from 12.8K to 488 tons (FAOStat). Cacao has been fundamental in the transformation of the rural economy dependent on illicit crops. As such, Cacao is called the "peace crop", nominated by  Colombia's President and 2016 Nobel peace prize winner, Juan Manuel Santos, and by the President of the Association of Cacao Growers (Fedecacao). Although Colombian cacao has the potential to be in the high value markets for fine flavour, the lack of expert support as well as the use of traditional, and often times suboptimal, technologies makes cocoa production negligibly.

\textcolor{blue}{This study reports the cacao model which was calibrated for five agroclimatic regions in Colombia. The model simulates crop development, growth and yield, and predicts the maturation day when the fruit is likely to be ready for harvest.  The purpose of this model was to help farmers achieve  higher quality cacao beans by providing them with a tool that not only predicts optimal harvest date but also estimated a potential yield and biomass . The low RRMSE for yield prediction indicates the simplicity of the Cacao model to simulated cacao  yield. Although, the original SIMPLE model can be used to asses heat and water stress, our cacao model calibration  did  not accounted biotic (pest and diseases) and abiotic (heat or water stress) factors that could have significant effects on the cacao production.  Santander and Caldas have the  highest solar radiation but the lowest temperature. In contrast , Arauca and Apartado have the  highest temperature but the lowest solar radiation
Weather analysis could not show a correlation between flower shedding and high wind or rain because information on flowers was not part of the field data collected. 	The growth cycle was simulated in terms of thermal time which was defined for the five regions tested, given the dependency to the crop model to predict the harvest day according to the weather variations while fruit are growing.  The thermal time calculation was defined base on  180 DAF because farmers cut the pods by calendar days. Our results showed how the harvest day can vary depending on the accumulate temperature during each specific crop cycle simulated (Fig. \ref{fig:dayH}).
}

@@

Our analysis of weather variables (Fig. \ref{fig:heat}) could not show a correlation between flower shedding and high wind or rain because information on flowers was not part of the field data collected. However, farmers stated that the number of flowers pollinated decrease by months where wind and rain are high. High winds can affect the availability of tiny flies pollinators from Diptera order and from the families of of the biting midges \textit{Ceratopogonidae},  genus  \textit{Forcipomyia} \citep{Saunders1959, kaufmann1975, sotomayor2020} to reach the cacao flowers. However, the stability of cacao flowers is influenced by seasonal wheather conditions (abiotic) and pollination (biotic) \citep{Frimpong2014}. Therefore, pollinator population should be coincidence with the phenology of the flowering cacao trees \citep{Young1983, Young2012}. Flower opening is very well synchronised between the cohorts of mature flowers opening each night \citep{Niemenak2010}. The flowers open at almost exactly the same time and rate, irrespective of their position on the trunk. Thus, unfertilised flowers abscise from the trunk approximately 1 day after flower opening  \citep{Niemenak2010}. Hence more than 90\% of unpollinated flowers fall or abscised within 32 hours after  anthesis \citep{Aneja1999}. Abscission processed of  flowers are mainly controlled  by  three  hormones: auxin, ethylene, and abscisic acid (ABA) \citep{Aneja1999}. Ethylene generally promotes abscission because it may  inhibit the transport of auxin from the leaf blade, which allow the action of ABA to promote the fall of flowers \citep{Beyer1975}. In general, environmental conditions can also  stimulate   to  a  decrease  the  auxin/ethylene relationship \citep{Aneja1999}.

When analysing the data from  Santanter region, we showed that the number of successful flowers pollinated to produce final yield could be affected by rain, TMAX and wind. Nevertheless, a better field data tracing flowers development is essential to understand if it is a mechanical or physiological effect.  \cite{Frimpong2009, Frimpong2011} indicated that the numbers of cocoa pollinators were reduced during the dry season, but increased in the wet season. This, could be due to midges need a moist environment to develop \citep{Frimpong2014}, which is difficult during the dry season as cocoa leaves create a dried ground mat \citep{Frimpong2009, Frimpong2014}. Moreover, the lack of water during dry seasons may reduce the nutrients uptake provoking the massive flower drops \citep{Vaughton2017}.  For future studies, wind could be included in the cocoa model as an input to simulate this mechanical effects over number of flowers. In general, field data regarding counting flowers pollinated by month should be better reported for the region here studied.

\subsection{Thermal Time for Harvest Day Predictions}
We define the thermal time required to harvest cocoa pods for five Colombian regions as maturation of the fruit is related with temperature during the growth cycle \citep{lopez2018}. Previous studies have calculated thermal time for  different cocoa cultivar in Brazil and Ghana \cite{Daymond2008}. They also confirm that the fruit maturation time decrease in with an increase in temperature as was presented on others researches \citep{Alvim1974, End1991, Daymond2008}. the effects of temperature and solar radiation on fruit growth and development was previously studied by \cite{Daymond2008}, showing that crop under higher temperatures thought the crop cycle induce greater fruit losses because of physiological maturation (cherelle wilt). When the fruit is mature, seeds are able to germinate  \citep{lopez2018}. \textcolor{blue}{However, when ripe fruits stay for a longer time on the tree without be harvested  on the right moment, seeds can germinate inside the pod damaging  cacao production for high-quality taste and affecting the final yield. }

@@

\subsection{Cocoa Crop Model Simulations}
Although cocoa is a relevant crop and there is an extensive agronomic literature, there is only one physiological crop model specific for cocoa so far. The (SUCROS-Cocoa) developed by \cite{zuidema2005}. However, the code was not easy available for adaptations. In contrast, the SIMPLE model has an open code in R, which we could adapt such a model would be very useful to compare yields and predict harvest date in different climates. As the harvest day was predicted from the FD (Fig. \ref{fig:dayH}, consequently, biomass production and fSolar presented a crop cycle shorter than 180 DAF (Fig. \ref{fig:m1}, a and b). These simulations are coincident with results presented by \cite{lopez2018}, where physiological maturity of coca pod varies from 140 to 162 DAF.  Our results showed how the harvest day can vary depending on the accumulate temperature during each specific crop cycle simulated (Fig. \ref{fig:dayH}). \textcolor{blue} {Harvesting cocoa fruits at the right moment the quality of the seed inside the pod can improve avoiding the germination before collect the pot from  cacao trees.}\textcolor{olive} {An research made by Gallego (2021), cacao clones CCN51 and ICS95 were handed-pollinated and pods were collected at 6, 7, and 8 months of ripeness after pollination. Cacao fruits were harvested from the municipality of Belalcazar, Caldas, Colombia to analyse metabolites related with cocoa quality were more expressed when fruits are collected at 6, 7, and 8 months after flowering.  Results showed that the metabolites that mostly explain the differences between the classes ( metabolites with the highest VIP) are highly overexpressed at month 6 and then decline or under-expressed at the next months.  No significant differences were found in their metabolic profile Between month 7 and 8.
Quality of cocoa is defined by  some metabolites that are associated with flavonoids or colouration. When the expression of flavonoids decrease , indicates  that  metabolites  associated with flavonoids  are being oxidized and metabolites are converted by cells into other less oxidizing metabolites. The colouration does not show a greater difference after month 6. If there is no good  accumulation of flavonoids which are also related with metabolites associated with colouration, the quality of cocoa seed decrease. Ripening is associated with changes in colouration and these in turn with flavonoids so month 6 was ideal for harvesting assuring the high quality of cocoa. } \textcolor{blue} { Therefore, the cacao model can predict with more detail the optimum day to harvest the pod ensuring the quality of the cacao bean  destined for Fine-cocoa products. }

@@

Yield prediction presented a RRMSE values 7.2\%, which were significant lower than those  presented for other crops using the SIMPLE model which reported and RRMSE of 24.4\% \citep{Zao2019simple}. Resulting in reliable approach for cocoa yield prediction.  \textcolor{blue}{The low RRMSE for yield prediction indicates the simplicity of the Cacao model to simulated cacao  yield. Although, the original SIMPLE model can be used to asses heat and water stress, our cacao model calibration  did  not accounted biotic (pest and diseases) and abiotic (heat or water stress) factors that could have significant effects on the cacao production.} In general, these results may help to improve the quality of cocoa seed considering the moment to harvest can be variable depending on weather changes.

@@

Cocoa fruit development for harvest in the right time depend on whether conditions and principles of crop physiology and flower phenology.  This was common for the five regions. Thermal time characterisation range from 1200 to 3000 days $^\circ$C, with a T$_{b}$ of 10 $^\circ$C for the fruit development.  The cocoa model allowed to predict the harvest date with better precision than only considering days by calendar. Thus, the crop cycle of cocoa for harvest  should be shorter than 180 days after flowering.

\textcolor{blue}{Results are only valid for the regions tested and they can be used for farmers in Colombia for improve the cacao quality.  However, the model calibration and values of parameters used may be applied by the international community to test cacao crops in other regions calculating the specific thermal time using the SIMPLE model. Moreover, other cacao varieties and regions to test can be in included. As for most of the crop modelling studies field data is extremely important .} These results confirm the potential of the Crop Simulation Model approaches  for tropical crops in Colombia and other regions in Latin America.
